---
title: "STAT 340 Final exam - Question 3"
author: "Kara Conrad"
date: "5/24/22"
output: html_document
---

```{r include=F}
knitr::opts_chunk$set(echo=T,warning=F,message=F,fig.align="center",error=T)
library(tidyverse)
library(lme4)
library(glmnet)
library(rvest)
```

***REMEMBER:***
 - Unless manual/by hand calculation specifically requested, you can **use any R functions covered in class**. Functions not covered in class MAY NOT be graded (depending on context).
 - **All plots MUST have reasonably good titles/labels** (quality of titles/labels subject to determination by graders). Failure to do so MAY result in penalties.
 - Also note that `error=TRUE` has been turned on in the setup chunk. This means that **the file will knit EVEN if there are errors**! This will ensure you always have an HTML file to submit, but remember to ***CHECK YOUR FILE FOR ERRORS BEFORE YOU SUBMIT!!*** You can turn this off if you like, but it seems to help most people.


## Question 3: Great Crested Newt
A survey of great crested newts was conducted in the UK in an effort to better understand their ecology and conserve their habitats.^[https://www.arguk.org/info-advice/advice-notes/9-great-crested-newt-habitat-suitability-index-arg-advice-note-5/file] Many ponds were carefully examined and checked for presence of newt life (0 indicating no newts were present; 1 indicating at least some newts were present). Several predictors were also recorded for each pond, listed below:
 - `log10area`: pond area (m²)
 - `dry`: how often the pond dries (1-4: never, rarely, sometimes, annually)
 - `water`: water quality (1-4: bad, poor, moderate, good)
 - `shade`: % of shoreline in shade
 - `bird`: evidence of bird impact (1-3: none, minor, major)
 - `fish`: evidence of fish impact (1-4: none, possible, minor, major)
 - `ponds`: # of ponds within 1km
 - `land`: land habitat quality (1-4: none, poor, moderate, good)
 - `macro`: % pond area covered by plants

Here's the dataset. We're going to explore different ways of predicting the presence of newts in ponds.
```{r}
newt = read.csv('newt.csv', header=TRUE)
head(newt)
```


### Part I <small>(pts: 1, 2, 2, 2)</small>
a. Create an 80/20 train/test split (i.e. randomly sample 80% of the rows to be used as a training dataset, and set aside the remaining 20% as a testing dataset).
```{r}
samp = sample(1:200, 40)
test = newt[samp,]
train = newt[-samp,]
```

b. Start by fitting a full logistic regression model (i.e. with all terms). **Show a summary table** of the resulting fit
```{r}
glm.default = glm(presence ~ ., data = train)
summary(glm.default)
```

c. Which predictors are significant? Write interpretations for the 3 most significant predictors. Be sure to indicate the following:
   - **the direction** of the relationship (i.e. is it linked to increasing or decreasing chance of newt presence?)
   - **the magnitude** of the relationship (i.e. how much are the chances increasing? Please use **precise** quantitative, statistical language).
 Ponds- positive direction, shade- negative direction, and fish- negative direction are the 3 most significant predictors
 ## cant fully answer
 
d. Use your fitted model to generate predictions on the test dataset and compare your predictions to the true test responses to **estimate the out-of-sample mean squared error** (MSE) of this model. **Report the MSE**. (*Hint:* remember to set the `type` argument appropriately to get actual probabilities).
```{r}
predict = predict(glm.default, test, type = "response")
data = data.frame(pred = predict, actual = test$presence)
mse = mean((data$actual - data$pred)^2)
mse
```



### Part II <small>(pts: 3, 1, 2)</small>
Let's try a different approach where we try to **simultaneously perform regularization and variable selection**.

a. Choose a regularization method that is also good at selecting out a smaller subset of variables (*Hint:* which method is good at setting coefficients to exactly 0?). Using the `glmnet` package, fit a model of this type, making sure to set the correct value for `alpha`. Also add the argument `family=binomial` to ensure you get a logistic-type fit.
```{r}
# create train x and y
x.train = model.matrix(presence ~ ., data = train )[,-1]
y.train = train$presence

# create test x and y
x.test = model.matrix(presence ~ ., data = test )[,-1]
y.test = test$presence

# get best lambda by using CV
lambda.best = cv.glmnet( x.train, y.train, alpha = 1 , family=binomial )$lambda.min
print(lambda.best)

# fit using best lambda
lasso.newt = glmnet( x.train, y.train, alpha = 1 , family=binomial , lambda=lambda.best )
```

b. Show the coefficients of your fit. Which predictors were removed from the model by your method? 
```{r}
# get coefficients
coef(lasso.newt)
```

c. Again, use your fitted model to generate predictions on the test dataset and compare your predictions to the true test responses to **estimate the out-of-sample mean squared error** (MSE) of this model. **Report the MSE** (*hint:* remember to set `type` appropriately!), and **compare this value to the MSE from the first model**. Is this MSE higher or lower? By what percent approximately?
```{r}
# get predictions
prediction = predict(lasso.newt, newx=x.test , type="response" )

# from here, can find the MSE
mse.lasso = mean( (prediction - y.test)^2 )
print(mse.lasso)

# report this MSE is higher/lower than the previous by what percent
percent.improvement = (mse.lasso - mse) / mse * 100
print(percent.improvement)
```
